Monitoring Time Series With Missing Values:\\a Deep Probabilistic Approach

### 2022-02-18 paper.tex
for submission

\begin{abstract}
Systems are commonly monitored for health and security through
collection and streaming of multivariate time series. Advances
in time series forecasting due to adoption of multilayer
recurrent neural network architectures make it possible to
forecast in high-dimensional time series, and identify and
classify novelties early, based on subtle changes in the trends.
However, mainstream approaches to multi-variate time series
predictions do not handle well cases when the ongoing forecast
must include uncertainty, nor they are robust to missing data.
We introduce a new architecture for time series monitoring based
on combination of state-of-the-art methods of forecasting in
high-dimensional time series with full probabilistic handling of
uncertainty. We demonstrate advantage of the architecture for
time series forecasting and novelty detection, in particular
with partially missing data, and empirically evaluate and
compare the architecture to state-of-the-art approaches on a
real-world data set.
\end{abstract}

@@

\begin{figure}[t]
\begin{subfigure}{.64\textwidth}
\begin{tikzpicture}
\node [rectangle, draw] (rnn) {RNN};

@@

A popular realization of the forecasting model
$\mathcal{M}_\theta$ is a recurrent neural network (RNN), with
$\theta$ corresponding to the network parameters.  There is a
range of neural recurrent models of varying complexity to deal
with time series forecasting. Most models include a recurrent
unit which threads the state through the time steps, accepts
data as inputs and produces next step predictions as outputs.
The simplest model is an RNN with a fully-connected readout
layer to produce forecasts
(Figure~\ref{fig:dt-conventional-model}).  RNN can be based on
LSTM~\cite{HS97}, GRU~\cite{CMB+14}, or another architectural variant, and
is often multi-layer.  Architectures  may also include
intermediate modules and sampling-based variational
layers~\cite{CKD+15,YB21}. The overall architecture stays almost
the same, with more connections, intermediate modules and
sampling-based variational layers.

\paragraph{Input and Output} This architecture normally accepts
observation vectors and outputs vectors of distribution
parameters for the belief distribution of the observations at
the next time step. In the case of frequently used normal
distributions, the output vector consists of means $\mu$ and
standard deviations $\sigma$ of all dimensions and is twice as
wide as the input.  This allows to also assess uncertainty
multiple steps into the future: one can repeatedly sample from
the observation belief distribution at each future step, and
feed the sample as the input to the following step.

\paragraph{Training} The model is trained to maximize
probability of prediction. In the most basic case, a single step
is predicted for each time step in the series. If a series has
10 steps, then steps $1 ... 9$ are the input, and the steps $2
... 10$ are the truth. Following~\eqref{eqn:theta-star}, the
network is trained to minimize negative log probability of true
observations given the predicted belief distributions.

\paragraph{Novelty detection} After training, the model is used
to produce predictions of future observations. There are two
related but different phenomena indication a novelty (anomaly)
in time series behavior:
\begin{enumerate}
\item Predicted volatility of the time series is high, that is, future
observations can only be forecast uncertainly (with high variance). 
\item Probability of actual observations, when observed, given a prediction
from a past state, is low.
\end{enumerate}
Either phenomenon, or both of them, can be used to alert about
novelties in the time series. In recurrent neural network
architectures, the hidden state ($h_t$ in
Figure~\ref{fig:dt-time-series-models}) can be used to identify
and classify anomalies.

However, the basic scheme outlined above poses difficulty in
applications with high-dimensional time series and partially
missing observations. Sampling based uncertainty assertion
impacts performance, and missing observations are often imputed
heuristically~\cite{LKW16,SYG+19}. An architecture which
incorporates confidence about data and in which observed and
predicted data are interchangeable is  highly desirable. For
example, if out of 5 components 3 were measured and 2 predicted
from an earlier step we want to input them both into the next
time step for further forecasting. In addition, the model
architecture should be capable of robust uncertainty prediction
and benefit from training with multiple steps of out-of-sample
data.

To achieve these objectives, \textbf{we propose the following
changes} (Figure~\ref{fig:dt-uncertainty-propagation}):

\begin{enumerate}
\item The input, as well as the output, is a vector of
distribution parameters. For simplicity, we will confine further
discussion to the case of independent normal belief
distributions for each component; however, other distribution
shapes can be used. For the independent normal
distributions, the distribution parameter vector consists of the
means followed by the standard deviations. If the data has 5
components, the input will be 10-dimensional. For observed data
--- measurements present at the current time step --- the
standard deviation is zero. For missing data the input is the
mean and the standard deviation as predicted from the preceding
time steps.

\item Since we introduce confidence into the input, we cannot
train the network myopically, on a single step prediction --- or
the standard deviations in the input data will always be zero,
and the network will never learn how to use them. To overcome
this, we train on multiple predicted steps. We feed each
prediction, without sampling, as input to the next step and
compute the loss as negative log probability of this number of
future points versus our prediction.
\end{enumerate}

@@

The ability of probabilistic forecasting with uncertainty, in
the form of multivariate normal distributions, far into the
future, opens opportunity for application to more robust
novelty detection approaches. Instead of detecting novelty based
on log probability of observations given predictions from the
past, which is prone to false positives due to observation
noise, novelties can be detected and analysed by comparing
predictions of the same time point from different points in the
past. In this case, KL-divergence between predictions provide a
theoretically sound and robust mechanism for detection of
anomalies, and is in particular relevant for monitoring of large
operation environments with high dimensionality of time series
and occasional missing values and heteroskedastic noise.

@@

We evaluate the proposed architecture on a data set of
monitoring a cluster of 100 computing nodes in the cloud. For
each node, the incoming and the outgoing network traffic (in
bytes) and the CPU usage (relative) are logged with 1 minute
resolution. 240 hours were logged, resulting in 12000 120-minute
3-dimensional samples.  We split the dataset into the training,
validation, and test as 80\%, 10\%, and 10\% correspondingly.
Since the original data set does not have any missing data
points, we emulated data sets with missing data by randomly
removing 5\%, 10\%, 20\%, and 50\% of the data.

\begin{table}[t]
\caption{Uncertainty propagation vs. `replace by the mean`.}
\label{tab:replace-by-mean}
	\setlength\tabcolsep{18pt}
	\centering
	\begin{tabular}{r|c c c c}
	 \textbf{missing} & \textbf{2} & \textbf{4} & \textbf{8} & \textbf{16} \\ \hline
		  \textbf{5\%} & 0.001 & 0.001 &  0.06 & 0.10  \\
		 \textbf{10\%} & 0.001 & 0.002 &  0.08 & 0.11  \\
		 \textbf{20\%} & 0.003 & 0.003 &  0.11 & 0.13  \\
		 \textbf{50\%} & 0.004 & 0.006 &  0.12 & 0.16  
	\end{tabular}
\end{table}
\begin{table}[t]
\caption{Uncertainty propagation vs. `replace by a random sample`.}
\vspace{1em}
\label{tab:replace-by-sample}
	\setlength\tabcolsep{18pt}
	\centering
	\begin{tabular}{r|c c c c}
	 \textbf{missing} & \textbf{2} & \textbf{4} & \textbf{8} & \textbf{16} \\ \hline
		  \textbf{5\%} & 0.04 & 0.04 & 0.13 & 0.14 \\
		 \textbf{10\%} & 0.06 & 0.06 & 0.16 & 0.24 \\
		 \textbf{20\%} & 0.11 & 0.12 & 0.20 & 0.27 \\
		 \textbf{50\%} & 0.18 & 0.19 & 0.28 & 0.30   
	\end{tabular}
\end{table}
We used a 3-layer GRU-based recurrent neural network with hidden
size 64 and 20\% dropout between layers. We trained the network
with lookahead depths (number of steps to forecast in the
future) 2, 4, 8, and 16 using the Adam optimizer with learning
rate 0.001, training for 20 epochs (sufficient for convergence).
We performed the training on a cloud computing node with 1
NVIDIA T4 GPU, 4 Intel Xeon Platinum CPUs, and 64 Gb memory. The
training of a single model took 20 minutes.

We compared our approach with conventional imputation methods
`replace by the mean' and `replace by a random sample'.
In the `replace by mean' method, a missing value is replaced 
by the mean of the forecast. In the `replace by a random
sample' method, a missing value is replaced by a random sample
drawn from the forecast. As a performance metrics, we
used per-point negative log-likelihood loss on the test set. 
Tables~\ref{tab:replace-by-mean} and~\ref{tab:replace-by-sample}
show the difference in loss between uncertainty propagation and
`replace by the mean' and `replace by a random sample',
correspondingly. The greater is the number, the worse is the
forecasting by each of the methods compared to uncertainty
propagation.
One can see that in all cases uncertainty propagation provides
better forecasts than either of the conventional methods.

\begin{figure}[t]
\includegraphics[width=0.9\linewidth]{dist-vs-mean.pdf}
\caption{Uncertainty propagation vs `replace by mean'. 95\% confidence intervals are shaded.}
\label{fig:dist-vs-mean}
\end{figure}
As an illustration of the advantage of uncertainty propagation,
consider Figure~\ref{fig:dist-vs-mean}, which shows forecasts
using uncertainty propagation and `replace by mean' in presence
of missing values. Forecasts through uncertainty propagation
result in adequate confidence intervals. However, when missing
values are replaced by the mean of the belief distribution,
further forecasts are overconfident and too many observations
fall outside of 95\% confidence intervals.

@@

We confined most of the discussion to the normal uncertainty shape.
Other distributions can be used instead of the normal
distributions where appropriate, provided their parameterization
allows to express a certain observation as well as an uncertain
belief. Analysis of distributions for representing uncertainty
and their feasible parameterization is a subject of ongoing
research. Another research direction worth exploring is
extension of the presented architecture to bidirectional
recurrent neural networks~\cite{BRH15}. Bidirectional RNNs allow
to account for both past and future observations where
appropriate, but make apparently make uncertainty propagation
more complicated. Still, preliminary results suggest that
uncertainty in bidirectional RNNs can be handled in a similar
manner, further facilitating efficient probabilistic uncertainty
propagation in a broader class of deep learning models for time
series.

>>>>

\begin{abstract}
Systems are commonly monitored for health and security through
collection and streaming of multivariate time series. Advances
in time series forecasting due to adoption of multilayer
recurrent neural network architectures make it possible to
forecast in high-dimensional time series, and identify and
classify novelties early, based on subtle changes in the trends.
However, mainstream approaches to multi-variate time series
predictions do not handle well cases when the ongoing forecasts
must include uncertainty, nor they are robust to missing data.
We introduce a new architecture for time series monitoring based
on combination of state-of-the-art methods of forecasting in
high-dimensional time series with full probabilistic handling of
uncertainty. We demonstrate advantage of the architecture for
time series forecasting and novelty detection, in particular
with partially missing data, and empirically evaluate and
compare the architecture to state-of-the-art approaches on a
real-world data set.
\end{abstract}

@@

We introduce here a recurrent neural network architecture which
facilitates uncertainty propagation. The architecture is capable
both of handling missing values and of multi-step forecasting.
We begin with description of conventional forecasting with RNNs.
Then, we describe our proposed architecture as an extension to
the conventional model.

\begin{figure}
\begin{subfigure}{.64\textwidth}
\begin{tikzpicture}
\node [rectangle, draw] (rnn) {RNN};

@@

\subsection{Conventional Forecasting}

A popular realization of the forecasting model
$\mathcal{M}_\theta$ is a recurrent neural network (RNN), with
$\theta$ corresponding to the network parameters.  There is a
range of neural recurrent models of varying complexity to deal
with time series forecasting. Most models include a recurrent
unit which threads the state through the time steps, accepts
data as inputs and produces next step predictions as outputs.
The simplest model is an RNN with a fully-connected readout
layer to produce forecasts
(Figure~\ref{fig:dt-conventional-model}).  RNN can be based on
LSTM~\cite{HS97}, GRU~\cite{CMB+14}, or another architectural variant, and
is often multi-layer.  Architectures  may also include
intermediate modules, and sampling-based variational
layers~\cite{CKD+15,YB21}. The overall architecture stays almost
the same, with more connections, intermediate modules and
sampling-based variational layers.

\paragraph{Input and Output} This architecture normally accepts
observation vectors and outputs vectors of distribution
parameters for the belief distribution of the observations at
the next time step. In the simplest case, the network produces a 
single output for each input, that is the dimensions of the
input and the output vector coincide. This corresponds to the
assumption of homoskedasticity of epistemic noise, and either
the mean squared error (corresponding to the Gaussian error
distribution) or the mean absolute error (corresponding to the
Laplace error distribution) is minimized.

More generally though, the epistemic noise is better modelled
heteroskedastically, and a two-parameter loss distribution,
with the location and the scale as the parameters.
In the case of the frequently used normal (Gaussian)
distribution, the output vector consists of means $\mu$
(location) and standard deviations $\sigma$ (scale) of all
dimensions and is twice as wide as the input.

\paragraph{Training} The model is trained to maximize
probability of prediction. In the most basic case, called
\textit{out-of-sample one-step} forecasting, a single step
is predicted for each time step in the series. In an $n$-step
time series, steps $1 ... n-1$ are used as the input, and steps
$1 ... n$ as the ground truth. Following~\eqref{eqn:theta-star}, the
network is trained to minimize negative log probability of true
observations given the predicted belief distributions. More
generally, a model can also be trained to predict more than a
single step at once into the future, however this is rarely used in
practice because the necessary size of the training data set
grows exponentially with the prediction depth. Instead, future
predictions are produced recurrently during forecasting.

\paragraph{Forecasting} Forecasting is accomplished by passing
past observations through the model to obtain forecasts for the
future time steps. In the \textit{out-of-sample one-step} mode,
a single step into the future is forecast. If a longer forecast
is required, the current forecast is entered as the input at the
next time step, time after time, up to the required length.
Either the location (the point forecast) or a random sample from
the belief distribution is used as the future input. Using
random samples also allows to assess uncertainty multiple steps
into the future: one can repeatedly sample from the belief
distribution at each future step, and feed the sample as the
input to the following step. Then, based on produced samples at
future steps, one can estimate uncertainty intervals. Such
Monte-Carlo handling of uncertainty is quite expensive
computationally though, because the standard deviation of
prediction error decreases as slowly as $\sqrt{N}$
with the number of samples $N$, on one hand, and uncertainty
may, in general, grow exponentially with prediction depth, on
the other hand.

\paragraph{Novelty detection} Forecasts produced by the model
can be used for a number of purposes, including decision making
and, in particular, novelty (anomaly) detection. There are two
related but different phenomena indicating a novelty
in time series behavior:
\begin{enumerate}
\item Predicted volatility of the time series is high, that is, future
observations can only be forecast uncertainly (with high variance). 
\item Probability of actual observations, when observed, given a prediction
from a past state, is low.
\end{enumerate}
Either phenomenon, or both of them, can be used to alert about
novelties in the time series. In recurrent neural network
architectures, the hidden state ($h_t$ in
Figure~\ref{fig:dt-time-series-models}) can be used to identify
and classify anomalies.

\subsection{Forecasting with Uncertainty Propagation}

The basic scheme outlined above poses difficulty in
applications with high-dimensional time series and partially
missing observations. Sampling based uncertainty assertion
impacts performance, and missing observations are often imputed
heuristically~\cite{LKW16,SYG+19}. An architecture which
incorporates confidence about data and in which observed and
predicted data are interchangeable is  highly desirable. For
example, if out of 5 components 3 were measured and 2 predicted
from an earlier step we want to input all of them into the next
time step for further forecasting. In addition, the model
architecture should be capable of robust uncertainty prediction
and benefit from training with multiple steps of out-of-sample
data.

Our proposed architecture is based on the observation that if
(at least) the location and the scale are used  to represent
forecasts, an observation (that is, certain knowledge at a given
step) can also be expressed using two parameters, by setting the
location to the observation, and the scale to 0. For the normal
distribution $\mathcal{N}(\mu, \sigma)$, the location and scale
parameterization is straightforward, corresponding to $\mu$ and
$\sigma$, however other belief distributions can be parameterized
by location and scale as easily, e.g. the log-normal,
Gamma, or Laplace distribution. For conciseness, we will confine
further discussion to the case of independent normal belief
distributions for each component; however, other distribution
shapes can also be used. Based on this observation,
\textbf{we propose the following
extension} to the conventional RNN-based forecasting model
(Figure~\ref{fig:dt-uncertainty-propagation}):

\begin{enumerate}
\item \textit{The input, as well as the output, is a vector of
distribution parameters}. For the independent normal
distributions, the distribution parameter vector consists of the
means followed by the standard deviations. If the data has 5
components, the input will be 10-dimensional. For observed data
--- measurements present at the current time step --- the
standard deviation is zero. For missing data the input is the
mean and the standard deviation as predicted from the preceding
time steps.

\item  Training can, in principle, be accomplished on data with
missing values, but training on data with missing values
incurs performance drawbacks and should be avoided. First,
handling missing values and replacing them with early
predictions introduces contingency in the forward run of the RNN
and slows down significantly the execution during training.
Second, missing values should, in general, themselves be viewed
as anomalies. One must be able to handle them during inference, 
but should not rely on their presence in the training data.

Therefore, we devise a scheme for \textit{training our model on data
that does not contain missing values}. Even in applications where
missing values are common in inference, training data without
missing values is usually readily available. However,
since we introduce confidence into the input, we cannot
train the network myopically, in out-of-sample one-step manner --- 
the standard deviations in the input data will always be zero,
and the network will never learn how to use them. To overcome
this, we train on multiple predicted steps. We feed each
prediction, without sampling, as input to the next step and
compute the loss as negative log probability of this number of
future points versus our prediction.
\end{enumerate}

@@

The ability of probabilistic forecasting with uncertainty, in
the form of multivariate normal distributions, far into the
future, opens opportunity for application to more robust novelty
detection approaches. Instead of detecting novelty based on log
probability of observations given predictions from the
past~\cite{CBK09}, which is prone to false positives due to
observation noise, novelties can be detected and analysed by
comparing predictions of the same time point from different
points in the past.  In this case,
KL-divergence between predictions provides a theoretically sound
and robust mechanism for detection of anomalies, and is in
particular relevant for monitoring of large operation
environments with high dimensionality of time series and
occasional missing values and heteroskedastic noise~\cite{ASH08,T19}.

@@

We evaluate the proposed architecture on a data set of
monitoring a cluster of 100 computing nodes in the cloud. For
each node, the incoming and the outgoing network traffic (in
bytes) and the CPU usage (relative) are logged with 1 minute
resolution. 240 hours were logged, resulting in 12000 120-minute
3-dimensional samples.  We split the dataset into the training,
validation, and test as 80\%, 10\%, and 10\% correspondingly.
Since the original data set does not have many missing data
points, we emulated data sets with missing data by randomly
removing 5\%, 10\%, 20\%, and 50\% of the data.

We used a 3-layer GRU-based recurrent neural network with hidden
size 64 and 20\% dropout between layers. We trained the network
with lookahead depths (number of steps to forecast in the
future) 2, 4, 8, and 16 using the Adam optimizer with learning
rate 0.001, training for 20 epochs (sufficient for convergence).
We performed the training on a cloud computing node with 1
NVIDIA T4 GPU, 4 Intel Xeon Platinum CPUs, and 64 Gb memory. The
training of a single model took 20 minutes.

\begin{table}
\caption{Uncertainty propagation vs. `replace by the mean`.}
\label{tab:replace-by-mean}
	\setlength\tabcolsep{18pt}
	\centering
	\begin{tabular}{r|c c c c}
	 \textbf{missing} & \textbf{2} & \textbf{4} & \textbf{8} & \textbf{16} \\ \hline
		  \textbf{5\%} & 0.001 & 0.001 &  0.06 & 0.10  \\
		 \textbf{10\%} & 0.001 & 0.002 &  0.08 & 0.11  \\
		 \textbf{20\%} & 0.003 & 0.003 &  0.11 & 0.13  \\
		 \textbf{50\%} & 0.004 & 0.006 &  0.12 & 0.16  
	\end{tabular}
\end{table}
\begin{table}
\caption{Uncertainty propagation vs. `replace by a random sample`.}
\vspace{1em}
\label{tab:replace-by-sample}
	\setlength\tabcolsep{18pt}
	\centering
	\begin{tabular}{r|c c c c}
	 \textbf{missing} & \textbf{2} & \textbf{4} & \textbf{8} & \textbf{16} \\ \hline
		  \textbf{5\%} & 0.04 & 0.04 & 0.13 & 0.14 \\
		 \textbf{10\%} & 0.06 & 0.06 & 0.16 & 0.24 \\
		 \textbf{20\%} & 0.11 & 0.12 & 0.20 & 0.27 \\
		 \textbf{50\%} & 0.18 & 0.19 & 0.28 & 0.30   
	\end{tabular}
\end{table}
We compared our approach with conventional imputation methods
`replace by the mean' and `replace by a random sample'.
In the `replace by the mean' method, a missing value is replaced 
by the mean of the forecast. In the `replace by a random
sample' method, a missing value is replaced by a random sample
drawn from the forecast. As a performance metrics, we
used per-point negative log-likelihood loss on the test set. 
Tables~\ref{tab:replace-by-mean} and~\ref{tab:replace-by-sample}
show the difference in loss between uncertainty propagation and
`replace by the mean' and `replace by a random sample',
correspondingly. The greater is the number, the worse is the
forecasting by each of the methods compared to uncertainty
propagation.
One can see that in all cases uncertainty propagation provides
better forecasts than either of the conventional methods.

\begin{figure}
\centering
\includegraphics[width=0.95\linewidth]{dist-vs-mean.pdf}
\caption{Uncertainty propagation vs `replace by the mean'. 95\% confidence intervals are shaded.}
\label{fig:dist-vs-mean}
\end{figure}
As an illustration of the advantage of uncertainty propagation,
consider Figure~\ref{fig:dist-vs-mean}, which shows forecasts
using uncertainty propagation and `replace by the mean' in presence
of missing values. Forecasts through uncertainty propagation
result in adequate confidence intervals. However, when missing
values are replaced by the mean of the belief distribution,
further forecasts are overconfident and too many observations
fall outside of 95\% confidence intervals.

@@

We confined most of the discussion to the normal uncertainty shape.
Other distributions can be used instead of the normal
distributions where appropriate, provided their parameterization
allows to express a certain observation as well as an uncertain
belief. Analysis of distributions for representing uncertainty
and their feasible parameterization is a subject of ongoing
research. Another research direction worth exploring is
extension of the presented architecture to bidirectional
recurrent neural networks~\cite{BRH15}. Bidirectional RNNs allow
to account for both past and future observations where
appropriate, but apparently make uncertainty propagation
more complicated. Still, preliminary results suggest that
uncertainty in bidirectional RNNs can be handled in a similar
manner, further facilitating efficient probabilistic uncertainty
propagation in a broader class of deep learning models for time
series.
